---
title: "In-Class: Space-Time Point Patterns"
subtitle: "BSTA 374 · Week 4 · 50 minutes · KDE slices, volume rendering, Knox and linked pairs"
author: "Eric M. Delmelle · Lehigh University"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "On this page"
    number-sections: false
    code-fold: show
    code-copy: true
    highlight-style: github
    fig-align: center
    embed-resources: true
execute:
  warning: false
  message: false
---

[qmd link](https://github.com/BSTA-374/space-timePointPattern/blob/main/index.qmd)

## The point of this session

Last class we asked whether a pattern was clustered. Today: whether it is
clustered *in time as well*, which is a different question with a different
answer and a different operational meaning.

A dataset pooled over six months hides everything that matters about an
outbreak. We will look at the same 330 cases six ways, and the six views do not agree ---
that disagreement is the lesson.

Data: `outbreak_class_sim.csv`, 330 simulated cases over 16 weeks in an
8 × 8 km study area. **The coordinates are already projected** (EPSG:32618,
metres), which is common when data come from a health department rather than a
geocoder.

```{r}
#| label: setup
#| code-fold: false

library(sf)
library(dplyr)
library(ggplot2)
library(spatstat)
library(plotly)

set.seed(374)

cases <- read.csv("outbreak_class_sim.csv") |>
  mutate(onset_date = as.Date(onset_date),
         day  = as.numeric(onset_date - min(onset_date)),
         week = day %/% 7 + 1)

win <- owin(c(0, 8000), c(0, 8000))
pp  <- ppp(cases$x_utm, cases$y_utm, window = win)

c(n = npoints(pp), per_km2 = intensity(pp) * 1e6, weeks = max(cases$week))
```

Note there is no `st_transform()` anywhere: the numbers are metres already. If
you pass them through `crs = 4326` by reflex you will put this city in the
Atlantic.

---

## 1. The two marginal views (5 min)

Look at space alone, then time alone. Both are honest; neither is enough.

```{r}
#| label: marginals
#| fig-width: 10
#| fig-height: 4

p1 <- ggplot(cases, aes(x_utm, y_utm)) +
  geom_point(size = 0.9, alpha = 0.5, colour = "firebrick") +
  coord_equal() + theme_minimal() +
  labs(title = "Space, pooled over 16 weeks", x = "x (m)", y = "y (m)")

p2 <- ggplot(cases, aes(week)) +
  geom_bar(fill = "grey40") + theme_minimal() +
  labs(title = "Time, pooled over the whole area", x = "week", y = "cases")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

**Question 1.** The map shows a diagonal band of cases and one dense patch. The
epidemic curve shows a rise and fall. Can you tell from these two plots whether
the band and the patch were active at the same time? Write one sentence before
running anything else.

---

## 2. Kernel density, and what the bandwidth decides (8 min)

```{r}
#| label: kde-bw
#| fig-width: 10
#| fig-height: 4

sigmas <- c(200, 600, 1500)
surf   <- lapply(sigmas, function(s) density(pp, sigma = s, eps = 50) * 1e6)
zmax   <- max(sapply(surf, max))

par(mfrow = c(1, 3), mar = c(1, 1, 2, 2))
for (i in seq_along(surf))
  plot(surf[[i]], main = paste0("sigma = ", sigmas[i], " m"),
       zlim = c(0, zmax), ribargs = list(las = 1))
```

Same data, same colour scale, three bandwidths. At 200 m the surface resolves
individual cases; at 1500 m the band and the patch merge into one blur. The
automatic selectors disagree with each other by design:

```{r}
#| label: selectors
#| code-fold: false

c(diggle = bw.diggle(pp), ppl = bw.ppl(pp), scott = mean(bw.scott(pp)))
```

**Question 2.** Which of the three surfaces would you put in front of a health
officer, and what would you tell them it does *not* show?

---

## 3. The same KDE, one slice per fortnight (8 min)

One bandwidth, one colour scale, eight time slices. This is a space-time kernel
estimate with the temporal smoothing done crudely --- a hard cut every two weeks.

```{r}
#| label: slices
#| fig-width: 11
#| fig-height: 6

cases$fortnight <- (cases$week - 1) %/% 2 + 1
labs <- paste("weeks", (sort(unique(cases$fortnight)) - 1) * 2 + 1, "-",
              sort(unique(cases$fortnight)) * 2)

slices <- lapply(sort(unique(cases$fortnight)), function(f) {
  sub <- cases[cases$fortnight == f, ]
  density(ppp(sub$x_utm, sub$y_utm, window = win), sigma = 600, eps = 50) * 1e6
})
zmax <- max(sapply(slices, max))

par(mfrow = c(2, 4), mar = c(1, 1, 2, 2))
for (i in seq_along(slices))
  plot(slices[[i]], main = labs[i], zlim = c(0, zmax), ribargs = list(las = 1))
```

Now the pooled map falls apart. The band was never a band: it is one hotspot in
different places at different times. The dense patch, by contrast, is in every
panel.

```{r}
#| label: centre-track
#| fig-width: 5.5
#| fig-height: 5.5

track <- cases |>
  group_by(week) |>
  summarise(x = mean(x_utm), y = mean(y_utm), n = n())

ggplot(track, aes(x, y)) +
  geom_path(colour = "grey60") +
  geom_point(aes(size = n, colour = week)) +
  scale_colour_viridis_c() + coord_equal() + theme_minimal() +
  labs(title = "Weekly mean centre of cases", x = "x (m)", y = "y (m)")
```

The mean centre migrates from the southwest to the northeast. It is a crude
summary --- one point per week, dragged around by the static patch --- but the
drift is unmistakable.

**Question 3.** Estimate the speed of the moving hotspot in metres per week off
the track plot or the cube below. What would you need to know about this disease to say whether
that speed is plausible?

---

## 4. The space-time cube (7 min)

Both marginal views and all eight slices in one object: easting, northing, and
time on the vertical axis. Rotate it.

```{r}
#| label: cube

plot_ly(cases, x = ~x_utm, y = ~y_utm, z = ~day,
        type = "scatter3d", mode = "markers",
        marker = list(size = 3, opacity = 0.7, color = ~day,
                      colorscale = "Viridis", showscale = TRUE)) |>
  layout(scene = list(xaxis = list(title = "x (m)"),
                      yaxis = list(title = "y (m)"),
                      zaxis = list(title = "day of study")))
```

Two structures should be visible once the cube turns: cases climbing diagonally
through the box, and a knot of cases standing at a fixed location across the full
height. Points, though, are hard to read through --- the front of the cloud hides
the back. Section 5 renders the same structure as a surface.

```{r}
#| label: animation

plot_ly(cases, x = ~x_utm, y = ~y_utm, frame = ~week,
        type = "scatter", mode = "markers",
        marker = list(size = 7, color = "firebrick", opacity = 0.75)) |>
  layout(xaxis = list(title = "x (m)", range = c(0, 8000)),
         yaxis = list(title = "y (m)", range = c(0, 8000)),
         title = "Cases by week")
```

---

## 5. Volume rendering: isosurfaces of the space-time density (8 min)

The scatter cube plots the cases, and points hide each other. What we want is the
*density* of cases in space-time --- a continuous field over $(x, y, t)$ --- and
the surfaces that enclose its high values. That is a 3D kernel estimate with two
bandwidths: `h` in metres and `lam` in days.

```{r}
#| label: st-density

gx <- seq(0, 8000, by = 200)
gy <- seq(0, 8000, by = 200)
gt <- seq(0, 112, by = 3)
grid <- expand.grid(x = gx, y = gy, t = gt)

h <- 700          # spatial bandwidth, m
lam <- 14         # temporal bandwidth, days

v <- numeric(nrow(grid))
for (i in seq_len(nrow(cases))) {
  v <- v + exp(-((grid$x - cases$x_utm[i])^2 + (grid$y - cases$y_utm[i])^2) / (2 * h^2)
               - (grid$t - cases$day[i])^2 / (2 * lam^2))
}
v <- v / (2 * pi * h^2) / (sqrt(2 * pi) * lam) * 1e6 * 7   # cases per km^2 per week

lv <- as.numeric(quantile(v, c(0.80, 0.92, 0.985)))
c(cells = length(v), max = max(v), levels = lv)
```

Three nested shells at those levels: an outer envelope, a middle surface, and an
opaque core. `caps` off so the volume is open at the box faces, and explicit
lighting --- without it the shells read as flat silhouettes rather than as
surfaces.

```{r}
#| label: isosurface

shell <- function(level, opacity, colour) {
  list(type = "isosurface",
       x = grid$x, y = grid$y, z = grid$t, value = v,
       isomin = level, isomax = max(v),
       surface = list(count = 1, fill = 1),
       caps = list(x = list(show = FALSE), y = list(show = FALSE),
                   z = list(show = FALSE)),
       opacity = opacity, showscale = FALSE,
       colorscale = list(list(0, colour), list(1, colour)),
       flatshading = FALSE,
       lighting = list(ambient = 0.55, diffuse = 0.85, specular = 0.25,
                       roughness = 0.5, fresnel = 0.2),
       lightposition = list(x = 6000, y = -9000, z = 400))
}

cols <- c("#4a7fb5", "#e8873a", "#a01a1a")
op   <- c(0.12, 0.28, 0.85)

fig <- plot_ly()
for (k in 1:3) fig <- do.call(add_trace, c(list(fig), shell(lv[k], op[k], cols[k])))

fig |>
  add_trace(data = cases, type = "scatter3d", mode = "markers",
            x = ~x_utm, y = ~y_utm, z = ~day, hoverinfo = "skip",
            marker = list(size = 1.4, color = "rgba(30,30,30,0.55)")) |>
  layout(showlegend = FALSE,
         scene = list(xaxis = list(title = "x (m)"),
                      yaxis = list(title = "y (m)"),
                      zaxis = list(title = "day of study"),
                      aspectmode = "manual",
                      aspectratio = list(x = 1, y = 1, z = 0.85),
                      camera = list(eye = list(x = 1.6, y = -1.7, z = 0.7))))
```

To drop the grey panes and grid lines behind the volume, add
`showbackground = FALSE, showgrid = FALSE` to each axis in `scene`. Keeping them
gives the eye something to judge position against; removing them looks cleaner.

The events themselves are drawn on top, as faint dust --- 1.4 px, 55% opaque.
They are almost invisible individually, which is the point: inside the shells
they crowd together, outside them they thin to a scatter, so the shells are
visibly *a summary of the points* rather than a shape drawn over them. It is the
quickest way to answer the question a student is right to ask, which is where the
cloud came from.

The structures now have surfaces. The outer envelope encloses an **inclined
tube** climbing from the southwest to the northeast --- the hotspot advancing
through the city --- and, standing beside it, a **vertical column** rooted at one
location and running the full height of the box. The core sits inside the tube
only: the travelling outbreak reaches intensities the endemic pocket never does.

For a cloud-like rendering of the same field, a single `volume` trace draws many
semi-transparent shells instead of three, with three orthogonal slice planes
through the interior:

```{r}
#| label: volume

plot_ly() |>
  add_trace(type = "volume",
            x = grid$x, y = grid$y, z = grid$t, value = v,
            isomin = as.numeric(quantile(v, 0.75)), isomax = max(v),
            opacity = 0.08, surface = list(count = 20),
            slices = list(z = list(show = TRUE, locations = c(30, 60, 90))),
            caps = list(x = list(show = FALSE), y = list(show = FALSE),
                        z = list(show = FALSE)),
            colorscale = "YlOrRd") |>
  layout(scene = list(aspectmode = "manual",
                      aspectratio = list(x = 1, y = 1, z = 0.85),
                      zaxis = list(title = "day of study")))
```

**Question 4.** You have now made three choices the picture depends on: `h`,
`lam`, and the levels. Re-render with the levels at the 95th and the 70th
percentile. At which one does the endemic column disappear, and what does that
tell you about reporting a hotspot as an object with a boundary?

::: {.callout-warning}
## Two practical limits
**Points inside an opaque shell.** The innermost shell is at opacity 0.85 rather
than 1 so the dust inside it still shows through. At 1 the core renders solid and
the densest events --- the ones that matter most --- disappear behind it.

**Grid size.** Each isosurface trace ships `x`, `y`, `z` and `value` to the
browser, so a 200 m / 3-day grid is already about 64,000 cells per trace and
three traces make a heavy page. Halving the spacing multiplies that by eight. Go
finer for a figure you export, not for a page you publish.

**plotly.js version.** `isosurface` and `volume` need a recent build; on an older
one the plot returns empty. The alternative is `misc3d::contour3d()` through
`rgl`, which renders smoother surfaces and is the better route for a
high-resolution still:

```r
A <- array(v, dim = c(length(gx), length(gy), length(gt)))
misc3d::contour3d(A, level = lv, x = gx, y = gy, z = gt,
                  color = cols, alpha = c(0.12, 0.3, 1), smooth = TRUE)
```
:::

---

## 6. Is there space-time interaction? The Knox test (7 min)

The cube is persuasive but it is not a test. The **Knox test** is the simplest
one: fix a distance threshold and a time threshold, count the pairs of cases that
are close on both, and compare that count with what you would get if the dates
were shuffled.

```{r}
#| label: knox
#| code-fold: false

d0 <- 600     # metres
t0 <- 14      # days

Ds <- as.matrix(dist(cbind(cases$x_utm, cases$y_utm))) < d0
Dt <- as.matrix(dist(cases$day)) < t0
diag(Ds) <- FALSE; diag(Dt) <- FALSE

knox_obs <- sum(Ds & Dt) / 2

perm <- replicate(499, {
  i <- sample(nrow(cases))
  sum(Ds & Dt[i, i]) / 2
})

c(observed = knox_obs, null_mean = mean(perm),
  p = (1 + sum(perm >= knox_obs)) / (1 + length(perm)))
```

```{r}
#| label: knox-plot
#| fig-height: 3.4

ggplot(data.frame(perm = perm), aes(perm)) +
  geom_histogram(bins = 30, fill = "grey80", colour = "white") +
  geom_vline(xintercept = knox_obs, colour = "firebrick", linewidth = 1) +
  theme_minimal() +
  labs(x = "close pairs under permuted dates", y = "",
       subtitle = "red = observed")
```

Permuting the dates destroys any link between where a case was and when, while
keeping both marginal distributions exactly as observed. That is the right null
for this question: it is a test of *interaction*, not of spatial clustering.

**Question 5.** The two thresholds were chosen by me. Change `d0` and `t0` --- try
300 m / 7 days and 1500 m / 42 days --- and re-run. Does the conclusion hold?
What does that tell you about a test that requires you to pick two numbers, and
which method from Wednesday avoids the choice?

---

## 7. Drawing the links (6 min)

The Knox count is a number; the pairs behind it are objects you can draw. Pick a
distance and a time threshold, join every pair of cases that is close on both,
and look at what the links connect.

```{r}
#| label: link-pairs
#| code-fold: false

d0 <- 600     # metres  -- change these two
t0 <- 14      # days

Ds <- as.matrix(dist(cbind(cases$x_utm, cases$y_utm))) < d0
Dt <- as.matrix(dist(cases$day)) < t0

pairs <- which(Ds & Dt & upper.tri(Ds), arr.ind = TRUE)
nrow(pairs)                                  # = the Knox count
```

```{r}
#| label: link-map
#| fig-width: 6
#| fig-height: 6

seg <- data.frame(x    = cases$x_utm[pairs[, 1]], y    = cases$y_utm[pairs[, 1]],
                  xend = cases$x_utm[pairs[, 2]], yend = cases$y_utm[pairs[, 2]])

ggplot() +
  geom_segment(data = seg, aes(x, y, xend = xend, yend = yend),
               colour = "steelblue", alpha = 0.25, linewidth = 0.3) +
  geom_point(data = cases, aes(x_utm, y_utm), colour = "firebrick",
             size = 0.8, alpha = 0.8) +
  coord_equal() + theme_minimal() +
  labs(title = paste0("Cases linked within ", d0, " m and ", t0, " days"),
       x = "x (m)", y = "y (m)")
```

The same links in the cube, where the slope of a link carries information the map
throws away --- a near-horizontal link joins two cases days apart in the same
place, a steep one joins cases weeks apart:

```{r}
#| label: link-cube

edges <- data.frame(
  x = as.vector(rbind(cases$x_utm[pairs[, 1]], cases$x_utm[pairs[, 2]], NA)),
  y = as.vector(rbind(cases$y_utm[pairs[, 1]], cases$y_utm[pairs[, 2]], NA)),
  z = as.vector(rbind(cases$day[pairs[, 1]],   cases$day[pairs[, 2]],   NA)))

plot_ly() |>
  add_trace(data = edges, type = "scatter3d", mode = "lines",
            x = ~x, y = ~y, z = ~z, hoverinfo = "skip",
            line = list(width = 1.2, color = "rgba(70,110,170,0.35)")) |>
  add_trace(data = cases, type = "scatter3d", mode = "markers",
            x = ~x_utm, y = ~y_utm, z = ~day,
            marker = list(size = 2.5, color = "firebrick", opacity = 0.8)) |>
  layout(showlegend = FALSE,
         scene = list(xaxis = list(title = "x (m)"),
                      yaxis = list(title = "y (m)"),
                      zaxis = list(title = "day of study"),
                      aspectmode = "manual",
                      aspectratio = list(x = 1, y = 1, z = 0.85)))
```

Chains of links form connected components, which is as close as this kind of data
gets to a candidate transmission chain:

```{r}
#| label: components
#| code-fold: false

g <- igraph::graph_from_edgelist(pairs, directed = FALSE)
g <- igraph::add_vertices(g, nrow(cases) - igraph::vcount(g))
comp <- igraph::components(g)

c(components = comp$no, largest = max(comp$csize),
  singletons = sum(comp$csize == 1))
```

At 600 m and 14 days expect two large components --- one strung along the
travelling outbreak, one sitting on the endemic pocket --- and a long tail of
isolated cases. Tighten to 300 m and 7 days and the large component fragments;
loosen to 1500 m and 42 days and the whole dataset collapses into a single
component, which tells you nothing.

**Question 6.** A link means two cases were close in space and close in time. Name
two reasons a linked pair might have nothing to do with each other, and one thing
you would need in order to claim transmission. Then decide whether you would show
this figure to a health officer at all.

::: {.callout-warning}
The picture is seductive and the inference is weak. Linkage graphs like this are
sometimes presented as transmission networks; they are a *thresholded proximity*
graph, and with a common source, a shared water tank or simply dense population
they will draw links between cases that never infected one another. The graph is
a hypothesis generator, and the two thresholds are yours, not the disease's.
:::

---

## Takeaways

1. Pooling over time turns a moving hotspot into a static band. Almost every
   published hotspot map has this problem somewhere in it.
2. Slice with one bandwidth and one colour scale, or you are comparing your own
   rendering choices rather than the data.
3. A vertical column in the cube and an inclined tube demand different
   responses: endemic transmission versus an outbreak with a direction. The
   pooled map shows neither.
4. An isosurface gives a hotspot a boundary, and that boundary is an artefact of
   three choices: `h`, `lam` and the level. Rotate and re-level before believing
   a shape.
5. Knox tests interaction, not clustering, because permuting dates preserves both
   marginals. It costs you two arbitrary thresholds; the space-time $K$ function
   is the version that scans over them.
6. Drawing the Knox pairs makes the statistic concrete and the over-claim easy: a
   proximity graph is not a transmission network.

Exercise 3 asks you to do all of this on a different dataset, plus the
space-time $K$ function itself.
